Impact of IPTp Regimen on Pregnancy Outcomes in a Malaria-Endemic Setting

### 2025-04-26 code/analysis-code/Statistical_Analysis.qmd
Update P

# Path management
library(here)              # For file path handling

# Data manipulation and wrangling
library(dplyr)             # Data manipulation
library(tidyverse)         # Core data science packages (includes ggplot2, dplyr, readr, etc.)
library(janitor)           # Clean column names
library(skimr)             # Data summary
library(lubridate)         # Date and time handling
library(forcats)           # Handling categorical variables

# Visualization
library(ggplot2)           # Data visualization (also part of tidyverse)
library(ggpubr)            # Publication-ready plots

# Tables and reporting
library(gtsummary)         # Summary statistics tables
library(gt)                # Table formatting
library(knitr)             # Report generation
library(kableExtra)        # Enhanced table output in markdown

# Analysis-specific packages
library(Amelia)            # Visualizing missing data
library(pwr)               # Power analysis
library(DiagrammeR)        # Diagrams and flowcharts
library(survival)          # Survival analysis
library(survminer)         # Survival plots
library(ggeffects)         # Marginal effects for regression models
library(broom)             # Tidy model outputs

# Machine learning and modeling
library(tidymodels)        # Unified modeling framework (includes recipes, parsnip, workflows, tune, etc.)
library(rsample)           # Resampling and data splitting
library(recipes)           # Preprocessing
library(parsnip)           # Model specification
library(workflows)         # Combine models and preprocessing
library(tune)              # Hyperparameter tuning
library(yardstick)         # Model performance metrics
library(dials)             # Parameter tuning grids
library(ranger)            # Random forest engine
library(xgboost)           # Boosting engine
library(themis)            # Handling class imbalance (e.g., SMOTE)

@@

# Check for multicollinearity using the VIF (Variance Inflation Factor)
library(car)
vif_values <- vif(model_interaction)
print(vif_values)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(here)

# Generate Analysis of Deviance Table (Likelihood Ratio Test)
anova(model_no_interaction, model_interaction, test = "LRT")
anova_df <- as.data.frame(anova_out)
anova_df$Model <- c("Model 1 (No Interaction)", "Model 2 (With Interaction)")
anova_df <- anova_df %>% 
  select(Model, `Resid. Df`, `Resid. Dev`, Df, Deviance, `Pr(>Chi)`)
colnames(anova_df) <- c("Model", "Residual Df", "Residual Dev", "Df", "Deviance", "p-value")

kable(
  anova_df,
  format = "pandoc",
  caption = "*Table X: Analysis of Deviance Comparing Models With and Without Interaction*"
) %>% 
  kable_styling(full_width = FALSE, position = "center")

@@

```{r}
library(dplyr)
library(knitr)
library(kableExtra)
library(car)
library(here)

# Compute Variance Inflation Factors (VIF) for Model 2 (with interaction)
vif_vals <- vif(model2)
vif_df <- as.data.frame(vif_vals)
vif_df$Variable <- rownames(vif_vals)
vif_df <- vif_df %>% rename(`Df (VIF)` = Df)
vif_df <- vif_df %>% 
  select(Variable, GVIF, `Df (VIF)`, `GVIF^(1/(2*Df))`) %>%
  mutate(
    Variable = case_when(
      Variable == "total_malaria_episodes" ~ "Total Malaria Episodes",
      Variable == "study_arm" ~ "Treatment Arm",
      Variable == "age_at_enrollment_years" ~ "Age at Enrollment",
      Variable == "gravidity" ~ "Gravidity",
      Variable == "education_level" ~ "Education Level",
      Variable == "total_malaria_episodes:study_arm" ~ "Interaction: Malaria Episodes × SP",
      TRUE ~ Variable
    )
  )

@@

@@

*Train/Test Split and Random Forest Model* which fits a random forest model and plots its ROC curve, is also used for Research Question 1 as an alternative ML approach

@@

```{r ml_improved_recipe, message=FALSE}
# 1) Encode factors, engineer features, balance classes, normalize
improved_recipe <- recipe(adverse_birth_outcome ~ 
                             total_malaria_episodes + study_arm +
                             age_at_enrollment_years + gravidity + education_level,
                           data = train_data) %>%
  # a) one‑hot encode all factor predictors
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  # b) create polynomial & interaction features
  step_mutate(
    malaria_sq       = total_malaria_episodes^2,
    malaria_DP_inter = total_malaria_episodes * study_arm_DP
  ) %>%
  # c) up‑sample minority class via SMOTE
  step_smote(adverse_birth_outcome) %>%
  # d) center & scale all predictors
  step_normalize(all_predictors())
                          # (5)

@@

@@

```{r}
# Create temporal & composite features in the training set
train_data <- train_data %>%
  mutate(
    # Month of enrollment (seasonality)
    enroll_month = lubridate::month(enrollment_date),
    # Episodes per trimester (assuming 40‑week gestation)
    episodes_per_trimester = total_malaria_episodes / (gestational_age_at_enrollment_weeks / (40/3)),
    # Socioeconomic index (if available)
    # wealth_index = as.numeric(education_level) * household_assets_score
  )

# Apply same transformations to test_data
test_data <- test_data %>%
  mutate(
    enroll_month = lubridate::month(enrollment_date),
    episodes_per_trimester = total_malaria_episodes / (gestational_age_at_enrollment_weeks / (40/3))
  )

@@

```{r}
imbalanced_recipe <- recipe(adverse_birth_outcome ~ ., data = train_data) %>%
  # standard preprocessing
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  # adaptive synthetic sampling
  step_adasyn(adverse_birth_outcome) %>%
  step_normalize(all_predictors())

@@

```{r}
# Continue using enet_wf from before
set.seed(2025)
enet_bayes <- tune_bayes(
  enet_wf,
  resamples = cv_folds,
  initial = enet_tune,       # seed the search from your grid results
  iter    = 20,              # number of Bayesian iterations
  metrics = metric_set(roc_auc)
)

@@

>>>>

# Path management
library(here)               # File path handling

# Data manipulation and wrangling
library(dplyr)              # Data manipulation
library(tidyverse)          # Core tidyverse packages (ggplot2, readr, etc.)
library(janitor)            # Clean column names
library(skimr)              # Quick data summaries
library(lubridate)          # Date/time handling
library(forcats)            # Categorical variable tools

# Visualization
library(ggplot2)            # Data visualization
library(ggpubr)             # Publication-ready plots

# Tables and reporting
library(gtsummary)          # Summary tables
library(gt)                 # Advanced table formatting
library(knitr)              # Report generation
library(kableExtra)         # Enhanced markdown tables

# Data exploration & preparation
library(Amelia)             # Missing-data visualization
library(pwr)                # Power analysis
library(DiagrammeR)         # Diagrams and flowcharts

# Survival analysis
library(survival)           # Survival models
library(survminer)          # Survival plots

# Model effects & outputs
library(ggeffects)          # Marginal effects extraction
library(broom)              # Tidy model outputs

# Machine learning & modeling
library(tidymodels)         # Modeling framework (recipes, parsnip, workflows, tune, yardstick, rsample)
library(themis)             # Class-imbalance sampling (e.g., SMOTE)
library(dials)              # Parameter tuning grids
library(ranger)             # Random forest engine
library(xgboost)            # Gradient boosting engine

@@

# Check for multicollinearity using the VIF (Variance Inflation Factor)
library(car)
vif_values <- vif(model_interaction)
print(vif_values)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Generate Analysis of Deviance Table (Likelihood Ratio Test)
anova_out <- anova(model_no_interaction, model_interaction, test = "LRT")

# Convert to a clean data frame
anova_df <- as.data.frame(anova_out)
anova_df$Model <- c("Model 1 (No Interaction)", "Model 2 (With Interaction)")

# Reorder and rename columns for readability
anova_df <- anova_df %>%
  select(Model, `Resid. Df`, `Resid. Dev`, Df, Deviance, `Pr(>Chi)`)

colnames(anova_df) <- c("Model", "Residual Df", "Residual Dev", "Df", "Deviance", "p-value")

# Display table
kable(
  anova_df,
  format = "pandoc",
  caption = "*Table X: Analysis of Deviance Comparing Models With and Without Interaction*"
) %>%
  kable_styling(full_width = FALSE, position = "center")

@@

```{r}
# Fit logistic regression model WITH interaction
model2 <- glm(adverse_birth_outcome ~ total_malaria_episodes * study_arm +
                age_at_enrollment_years + gravidity + education_level,
              family = binomial(link = "logit"),
              data = promo_data_clean)

@@

```{r}
# Compute Variance Inflation Factors (VIF) for Model 2 (with interaction)
vif_vals <- vif(model2)
vif_df <- as.data.frame(vif_vals)
vif_df$Variable <- rownames(vif_vals)
vif_df <- vif_df %>% rename(`Df (VIF)` = Df)
vif_df <- vif_df %>% 
  select(Variable, GVIF, `Df (VIF)`, `GVIF^(1/(2*Df))`) %>%
  mutate(
    Variable = case_when(
      Variable == "total_malaria_episodes" ~ "Total Malaria Episodes",
      Variable == "study_arm" ~ "Treatment Arm",
      Variable == "age_at_enrollment_years" ~ "Age at Enrollment",
      Variable == "gravidity" ~ "Gravidity",
      Variable == "education_level" ~ "Education Level",
      Variable == "total_malaria_episodes:study_arm" ~ "Interaction: Malaria Episodes × SP",
      TRUE ~ Variable
    )
  )

@@

#  Evaluating Discrimination (ROC Curve and AUC)

@@

*Train/Test Split and Random Forest Model* which fits a random forest model and plots its ROC curve, is also used for Research Question 1 as an alternative ML approach
```{r}
# 1. Define Random Forest model
rf_model <- rand_forest(
  trees = 500,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# 2. Build the recipe (you already did this correctly)
rf_recipe <- recipe(adverse_birth_outcome ~ total_malaria_episodes + study_arm +
                      age_at_enrollment_years + gravidity + education_level,
                    data = train_data) %>%
  step_dummy(all_nominal_predictors())

# 3. Combine into a workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)

# 4. Tune hyperparameters (optional next step)
set.seed(123)
rf_tune <- tune_grid(
  rf_workflow,
  resamples = vfold_cv(train_data, v = 10, strata = adverse_birth_outcome),
  grid = 20,  # number of hyperparameter combos
  metrics = metric_set(roc_auc)
)

@@

```{r ml_improved_recipe, message=FALSE, warning=FALSE}
# 0) Cast to correct classes *before* creating cv_folds
train_data <- train_data %>%
  mutate(
    total_malaria_episodes = as.numeric(total_malaria_episodes),
    gravidity              = as.numeric(gravidity),
    adverse_birth_outcome  = factor(adverse_birth_outcome, levels = c(0,1))
  )

test_data <- test_data %>%
  mutate(
    total_malaria_episodes = as.numeric(total_malaria_episodes),
    gravidity              = as.numeric(gravidity),
    adverse_birth_outcome  = factor(adverse_birth_outcome, levels = c(0,1))
  )

# 0b) Now create your 10-fold resamples on the cleaned train_data
cv_folds <- vfold_cv(train_data, v = 10, strata = adverse_birth_outcome)

# 1) Build the recipe: dummies, features, SMOTE, normalize
improved_recipe <- recipe(adverse_birth_outcome ~ 
                             total_malaria_episodes +
                             study_arm +
                             age_at_enrollment_years +
                             gravidity +
                             education_level,
                           data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_mutate(
    malaria_sq       = total_malaria_episodes^2,
    malaria_DP_inter = total_malaria_episodes * study_arm_DP
  ) %>%
  step_smote(adverse_birth_outcome) %>%
  step_normalize(all_numeric_predictors())

# 2) Specify elastic-net logistic regression
enet_spec <- logistic_reg(
    penalty = tune(),
    mixture = tune()
  ) %>%
  set_engine("glmnet")

# 3) Assemble into a workflow
enet_wf <- workflow() %>%
  add_recipe(improved_recipe) %>%
  add_model(enet_spec)

# 4) Tuning grid & 10-fold CV
enet_grid <- grid_regular(
  penalty(range = c(-5, -1), trans = log10_trans()),  # from 1e-5 to 1e-1
  mixture(range = c(0, 1)),
  levels = 5
)

# 5) Show the top AUC
enet_tune %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean))
```

```{r}
library(dplyr)
library(knitr)

enet_tune %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>%
  slice_head(n = 5) %>%                # ← pull the top 5 rows
  mutate(`AUC (mean ± SE)` = sprintf("%.3f \u00B1 %.3f", mean, std_err)) %>%
  select(penalty, mixture, `AUC (mean ± SE)`) %>%
  kable(
    caption = "Table X. Top 5 elastic‐net hyperparameter combinations, ranked by mean cross‐validated AUC"
  )

@@

```{r}
enet_tune %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>%
  head(5) %>%                          # ← base‐R head()
  mutate(`AUC (mean ± SE)` = sprintf("%.3f \u00B1 %.3f", mean, std_err)) %>%
  select(penalty, mixture, `AUC (mean ± SE)`) %>%
  kable(
    caption = "Table X. Top 5 elastic‐net hyperparameter combinations, ranked by mean cross‐validated AUC"
  )

```{r}
library(ggplot2)

enet_tune %>% 
  collect_metrics() %>% 
  filter(.metric=="roc_auc") %>% 
  ggplot(aes(x = log10(penalty), y = mixture, fill = mean)) +
    geom_tile() +
    scale_fill_viridis_c(name = "AUC") +
    labs(
      x = expression(log[10](lambda)),
      y = expression(alpha),
      title = "Heatmap of cross-validated AUC\nfor Elastic-net hyperparameter grid"
    ) +
    theme_minimal(base_size = 12)

@@

```{r}
# Create temporal & composite features in the training set
train_data <- train_data %>%
  mutate(
    enroll_month = lubridate::month(enrollment_date),
    episodes_per_trimester = total_malaria_episodes / (gestational_age_at_enrollment_weeks / (40/3))
  )

test_data <- test_data %>%
  mutate(
    enroll_month = lubridate::month(enrollment_date),
    episodes_per_trimester = total_malaria_episodes / (gestational_age_at_enrollment_weeks / (40/3))
  )

@@

```{r}
imbalanced_recipe <- recipe(adverse_birth_outcome ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_mutate(
    total_malaria_episodes = as.numeric(total_malaria_episodes),
    gravidity = as.numeric(gravidity)
  ) %>%
  step_adasyn(adverse_birth_outcome) %>%
  step_normalize(all_predictors())

@@

```{r}
enet_wf <- workflow() %>%
  add_model(enet_spec) %>%
  add_recipe(imbalanced_recipe)

enet_bayes <- tune_bayes(
  enet_wf,
  resamples = cv_folds,
  initial = enet_tune,
  iter = 20,
  metrics = metric_set(roc_auc)
)

@@

# More Machine Learning Analysis
To explore whether flexible, data‐driven classifiers could outperform our regression models in predicting the composite adverse birth outcome, we built a standardized ML pipeline using **tidymodels**. We compared three algorithms—LASSO‐penalized logistic regression, random forest, and XGBoost—each trained on the same predictors: maternal age, gravidity, education level, total malaria episodes, treatment arm, and socioeconomic status.

```{r}
# 2. Read & split data
df <- read_csv(here("data","clean","PROMO_Data_clean.csv")) %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1)))

set.seed(2025)
split    <- initial_split(df, prop = 0.7, strata = adverse_birth_outcome)
train    <- training(split)
test     <- testing(split)
cv_folds <- vfold_cv(train, v = 5, strata = adverse_birth_outcome)

```{r}
# 3. Preprocessing recipe
ml_rec <- recipe(adverse_birth_outcome ~ age_at_enrollment_years
                  + gravidity
                  + total_malaria_episodes
                  + education_level
                  + study_arm
                  + socioeconomic_status,
                data = train) %>%
  step_dummy(all_nominal_predictors()) %>%  # one‐hot encode
  step_zv(all_predictors()) %>%             # remove zero‐variance
  step_normalize(all_numeric_predictors())  # center & scale

```{r}
# 4. Model specifications & workflows

## (a) LASSO‐penalized logistic regression
lasso_mod <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

lasso_wf  <- workflow() %>% add_model(lasso_mod) %>% add_recipe(ml_rec)

## (b) Random forest
rf_mod <- rand_forest(mtry = tune(), trees = 1000, min_n = tune()) %>%
  set_engine("ranger") %>%
  set_mode("classification")

rf_wf  <- workflow() %>% add_model(rf_mod) %>% add_recipe(ml_rec)

## (c) XGBoost
xgb_mod <- boost_tree(trees = tune(),
                      tree_depth = tune(),
                      learn_rate = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

xgb_wf <- workflow() %>% add_model(xgb_mod) %>% add_recipe(ml_rec)

```{r}
# 5. Tuning grids & 5‐fold CV
set.seed(2025)

lasso_grid <- grid_regular(penalty(), levels = 20)

rf_params <- parameters(rf_mod) %>%
  update(mtry = mtry(c(1, 6)))
rf_grid   <- grid_latin_hypercube(rf_params, size = 20)

xgb_params <- parameters(xgb_mod)
xgb_grid   <- grid_latin_hypercube(xgb_params, size = 30)

ml_metrics <- metric_set(roc_auc)

lasso_res <- tune_grid(lasso_wf, resamples = cv_folds,
                       grid = lasso_grid, metrics = ml_metrics)

rf_res    <- tune_grid(rf_wf,    resamples = cv_folds,
                       grid = rf_grid,    metrics = ml_metrics)

xgb_res   <- tune_grid(xgb_wf,   resamples = cv_folds,
                       grid = xgb_grid,   metrics = ml_metrics)

```{r}
# 6. Collect cross‐validated AUC results
lasso_auc <- lasso_res %>% filter(.metric=="roc_auc") %>% summarize(mean = mean(mean), se = mean(std_err)) %>% mutate(model="LASSO")
rf_auc    <- rf_res    %>% filter(.metric=="roc_auc") %>% summarize(mean = mean(mean), se = mean(std_err)) %>% mutate(model="RF")
xgb_auc   <- xgb_res   %>% filter(.metric=="roc_auc") %>% summarize(mean = mean(mean), se = mean(std_err)) %>% mutate(model="XGB")

bind_rows(lasso_auc, rf_auc, xgb_auc) %>%
  rename(AUC = mean, SE = se) %>%
  knitr::kable(digits=3, caption="Table 6: 5‐fold CV ROC AUC for Each Model")

### 2025-04-26 products/manuscript/Manuscript.qmd
Update3
Update P

When evaluated on the independent test set, all four approaches performed at or below chance: logistic regression AUC = 0.48, random forest = 0.42, XGBoost = 0.44, and elastic net = 0.49 (Table 6; Figure 4). Even the strongest elastic net model fell essentially to random levels, underscoring that these routinely collected demographic and clinical variables lack sufficient predictive signal for reliable individual risk classification. These results highlight the need to integrate richer clinical, laboratory, or biomarker data in future predictive efforts to achieve meaningful discrimination.

>>>>

The best elastic-net configuration (penalty = 0.01, mixture = 0) achieved a cross-validated AUC of 0.606 on the training folds. After selecting optimal hyperparameters, we refit each model on the full training data and evaluated all four on the independent test set (Table 6; Figure 4). Discrimination was uniformly poor: logistic regression AUC = 0.48, random forest = 0.42, XGBoost = 0.44, and elastic net = 0.49—essentially at chance levels. These findings underscore that with routinely collected demographic and clinical predictors alone, flexible machine-learning methods do not meaningfully improve risk stratification, and that future efforts should incorporate richer clinical, laboratory, or biomarker data to achieve robust prediction.
